Real-time Attention Assessment for Social Robots

### 2015-09-14 paper.tex
Wrapped the whole paper at 80 cols
This will ease diffing

\maketitle
\begin{abstract}
Successfully sustaining face-to-face interaction between humans and robots mandates fine perception of each other attentional focus. Along this line, we present a novel technique to assess the visual focus of attention (VFOA) based on accurate, real-time head pose estimation with monocular camera. The software is fully open-source and can readily be adopted on a wide range of social robots. We propose an in-depth evaluation of the technique, using manual video annotations as ground truth, and show that our system reaches high level of accuracy in a real-world face-to-face interaction scenario. We finally show how to derive a metric of engagement from the monitoring of the VFOA, allowing the robot to account and react to the user's changing engagement in real-time.
\end{abstract}

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.9\columnwidth]{system}
    \caption{\small System overview showing the most relevant components involved in the estimation.}
    \label{system}
\end{figure}

@@

\section{Case Study}
The presented method have been successfully employed in a real child-robot interaction situation where the robot tries to engage the child in handwriting tasks using learning by teaching paradigm.
\subsection{Experimental Set-up}

@@

\begin{figure*}
    \centering
    \includegraphics[width=1.8\columnwidth]{realExpected}
    \caption{\small In blue, real VFOA of subject 4 accoding to two observer's assessment. In cyan, the matching between the real and the obtained VFOA. In yellow, the repeated pattern in turn-taking and in orange, the experimenter's suggestion to choose a new word.}
    \label{fig:realExpected}

@@

\begin{figure*}
    \centering
    \includegraphics[width=1.81\columnwidth]{realCaptured}
    \caption{\small Real VFOA of subject 4 accoding to two observer's assessment. In green, the obtained VFOA using our aproach and in cyan, the mathing between them. Finally, the correspondent numbers with figure \ref{fig:bitmap}.}
    \label{fig:realCaptured}
\end{figure*}

@@

Once the turn-taking is established, a pattern becomes visible during the subject's feedback, when an attempt is performed: Such pattern shows a high intermittent gaze frequency from the tablet to the selection. It also reveals that the subject is using the demonstration model in the selection tablet to provide a better answer to the robot. In fact, subjects decrease the response time after each shift as well as the gaze frequency towards the model suggesting an assimilation over time. Moreover, when a new demonstration is selected for playing, the same behavior is replicated (see yellow boxes in figure \ref{fig:realExpected}).
\\
\subsection{Focus of attention distribution}
Additional information of the head pose can be obtained for the evaluation of the VFOA distribution. Figure \ref{fig:heatmap} shows the most salient focuses of attention where the peaks are located. These peaks are mainly projected between the triangle composed by the robot, the writing tablet and the selection tablet using an unsupervised K-means clustering method reveals, considering the observers as negligible VFOA.

@@

\subsection{Engagement assessment}
If we consider the case shown in section \ref{fromFaceTo}, evaluating the subject's progression together with the time on task may be an indicator of engagement: If the subject answers so quickly with respect to the previous attempts with a poorer handwriting may be the case she is not engaged anymore. In such hypothetical case an activity switch can be easily implemented in a robot architecture. It confirms the findings reported in \cite{anzalone} where time response to event is proposed to account quality of interaction.

@@

%ACKNOWLEDGMENTS are optional
\section*{Acknowledgments}
This research was partially supported by the Funda\c{c}\~{a}o para a Ci\^{e}ncia
e a Tecnologia (FCT) with reference UID/CEC/ 50021/2013, and by the Swiss
National Science Foundation through the National Centre of Competence in
Research Robotics.

>>>>

\maketitle
\begin{abstract}

Successfully sustaining face-to-face interaction between humans and robots
mandates fine perception of each other attentional focus. Along this line, we
present a novel technique to assess the visual focus of attention (VFOA) based
on accurate, real-time head pose estimation with monocular camera. The software
is fully open-source and can readily be adopted on a wide range of social
robots. We propose an in-depth evaluation of the technique, using manual video
annotations as ground truth, and show that our system reaches high level of
accuracy in a real-world face-to-face interaction scenario. We finally show how
to derive a metric of engagement from the monitoring of the VFOA, allowing the
robot to account and react to the user's changing engagement in real-time.

\end{abstract}

@@

\begin{figure}[h!]
    \centering
    %\includegraphics[width=0.9\columnwidth]{system}
    \caption{\small System overview showing the most relevant components involved in the estimation.}
    \label{system}
\end{figure}

@@

\section{Case Study}

The presented method have been successfully employed in a real child-robot
interaction situation where the robot tries to engage the child in handwriting
tasks using learning by teaching paradigm.

\subsection{Experimental Set-up}

@@

\begin{figure*}
    \centering
    \includegraphics[width=1.8\columnwidth]{realExpected}
    \caption{\small In blue, real VFOA of subject 4 accoding to two observer's
    assessment. In cyan, the matching between the real and the obtained VFOA. In
    yellow, the repeated pattern in turn-taking and in orange, the experimenter's
    suggestion to choose a new word.}

    \label{fig:realExpected}

@@

\begin{figure*}
    \centering
    \includegraphics[width=1.81\columnwidth]{realCaptured}
    \caption{\small Real VFOA of subject 4 accoding to two observer's
    assessment. In green, the obtained VFOA using our aproach and in cyan, the
    mathing between them. Finally, the correspondent numbers with figure \ref{fig:bitmap}.}

    \label{fig:realCaptured}
\end{figure*}

@@

Once the turn-taking is established, a pattern becomes visible during the
subject's feedback, when an attempt is performed: Such pattern shows a high
intermittent gaze frequency from the tablet to the selection. It also reveals
that the subject is using the demonstration model in the selection tablet to
provide a better answer to the robot. In fact, subjects decrease the response
time after each shift as well as the gaze frequency towards the model suggesting
an assimilation over time. Moreover, when a new demonstration is selected for
playing, the same behavior is replicated (see yellow boxes in figure
\ref{fig:realExpected}).

\subsection{Focus of attention distribution}

Additional information of the head pose can be obtained for the evaluation of
the VFOA distribution. Figure \ref{fig:heatmap} shows the most salient focuses
of attention where the peaks are located. These peaks are mainly projected
between the triangle composed by the robot, the writing tablet and the selection
tablet using an unsupervised K-means clustering method reveals, considering the
observers as negligible VFOA.

@@

\subsection{Engagement assessment}

If we consider the case shown in section \ref{fromFaceTo}, evaluating the
subject's progression together with the time on task may be an indicator of
engagement: If the subject answers so quickly with respect to the previous
attempts with a poorer handwriting may be the case she is not engaged anymore.
In such hypothetical case an activity switch can be easily implemented in a
robot architecture. It confirms the findings reported in \cite{anzalone} where
time response to event is proposed to account quality of interaction.

@@

%ACKNOWLEDGMENTS are optional
\section*{Acknowledgments}

This research was partially supported by the Funda\c{c}\~{a}o para a Ci\^{e}ncia
e a Tecnologia (FCT) with reference UID/CEC/ 50021/2013, and by the Swiss
National Science Foundation through the National Centre of Competence in
Research Robotics.
